%%=============================================================%%
%% Back matter — npj / Springer Nature declarations
%%=============================================================%%
\backmatter

\bmhead{Data availability}
The NCT-CRC-HE-100K and CRC-VAL-HE-7K datasets used in this study are publicly available histopathology benchmarks (Kather et al.). No new datasets were generated in this study. \textcolor{red}{[Authors: insert the exact repository DOI/URL from which the datasets were obtained before submission.]}

\bmhead{Code availability}
The source code for DCA is publicly available at \url{https://github.com/xxx/DCA} to facilitate reproducibility and further research.

\bmhead{Acknowledgements}
The authors thank the reviewers for their valuable feedback that helped improve the quality of this manuscript.

\section*{Declarations}

\begin{itemize}
\item \textbf{Funding.} Not applicable.
\item \textbf{Competing interests.} The authors declare no competing interests.
\item \textbf{Ethics approval and consent to participate.} Not applicable; this study uses publicly available, de-identified datasets.
\item \textbf{Consent for publication.} Not applicable.
\item \textbf{Author contribution.} All authors contributed to the study conception, methodology, experiments, and manuscript preparation, and approved the final manuscript.
\item \textbf{Use of generative AI.} Portions of the text, including language polishing and editing, were assisted by a generative AI tool. The authors have reviewed and verified all content to ensure accuracy, originality, and compliance with scientific and ethical standards. No data analysis, results generation, or intellectual contributions were made by AI tools.
\end{itemize}

%%=============================================================%%
%% References. The .bbl produced from references.bib is inlined
%% at acceptance stage; for compilation we use the bib file.
%%=============================================================%%
\bibliography{references}

\end{document}
